### 2016-10-06 oldmort.Rnw
add reference

\title{Social inequality in mortality among
  adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
    at the
    session 'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, 
    Chicago, Illinois
    November 17--20, 2016. (Version 2016-10-03).}}

@@

\begin{abstract}

@@

{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}

\subsection{Combining the data sets and tabulate}

<<readtab>>=
load("data/vbTab.rda")
@

\section{Results}

@@

@@

{\huge Här ska 'conclusion.tex' in !!!}

>>>>

\title{Social inequality in mortality among
  adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
    at the
    session 'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, 
    Chicago, Illinois
    November 17--20, 2016. (Version 2016-10-06).}}

@@

\begin{abstract}
A long-term perspective on the development of social
inequalities in mortality in the adult and elderly population is taken. 
The area under investigation consists of the Skellefteå and Umeå regions in
the north of Sweden, and the time period is 1851--2013.

The main findings are that the social gradient becomes evident in the later
alf of the 20th century, with lower classes having the highest mortality.

@@

%%{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}

%%\subsection{Combining the data sets and tabulate}

<<readtab>>=
load("data/vbTab.rda")
@ 
\section{Introduction}

A long-term perspective on the development of social
inequalities in mortality in the adult and elderly population is taken. 
The area under investigation consists of the Skellefteå and Umeå regions in
the north of Sweden, and the time period is 1851--2013 for Skellefteå and
1901--2013 for Umeå.

This paper is a follow-up of an earlier paper \citep{segb16b}, presented at
the ESHD Conference in Leuven, Belgium, 2016. Here we make quick look back
in time for the Skellefteå region (1851--1900), and we also consider causes
of death for the later part of the 20th century. The {\sc Lisa} data allow
us to study the effect of income and education on mortality in a more
direct way.

\section{Aims and questions}

The aim of the paper is to investigate the development of social 
differences using HISCLASS in mortality for the adult and elderly
population during the mortality 
transition and its relation to the development of income inequality. The
analyses 
will be performed on data from the Skellefteå and Umeå regions
1901--2013. The questions are:

\begin{enumerate}
\item Has inequality in mortality between social classes increased among the adult and elderly
  population, primarily benefiting the highest social positions?

%%\item Is the mortality transition initiated by a certain social class?

\item Are there any gender differences in the effect of social position?

\item Is social position equally important among the retired population as for those in working age?
\end{enumerate}

\section{Previous research}

For a detailed presentation we refer to \citet{segb16}. Her is a short
summary.

Several studies have shown how important access to economic and other
resources are for health, but it is not clear how this has developed in
history. \citet{ant67} and \citet{smi83} suggest that social inequality in
health has passed through different phases in history. According to them,
social health differences were
comparatively small during the pre-transitional phase when space was a
strong determinant for the spread of disease. Differences then increased
during the transitional phase when mortality declined and wealthy groups
used their resources to gain better health. Finally, health differences
decreased again in modern low-mortality settings when instead
health-related behaviour became important, resulting in marginal
differences. \citet{omran82} state, in the third proposition in his theory
of the epidemiological transition, that even if the class differential in
mortality was maintained during the transition, the decline
set in earlier and was faster among privileged groups.

The present state of knowledge about the development of social 
health inequalities (in Sweden but also in other European countries), no
clear social hierarchy has been found before 1900 but substantial
differences towards the late 20th and early 21st century. It appears as if
a more consequent social health divide developed in the period
in-between. This period has not been thoroughly studied on this issue,
mainly because of lack of data to analyse. However, the extension of data
at the Demographic Data Base, Umeå University (DDB) makes such studies
possible. We are now able to analyse social 
differences in mortality with micro-data on a sufficiently large population
and with substantial social diversity from the 20th century for the
Skellefteå and Umeå regions \citep{wee16}.

\section{Data and variables}

The data for the present study comes from two large population databases at
the Demographic Data Base, Umeå University. The early period is covered by
the database Poplink, the digitisation of historical parish registers for
the Skellefteå and Umeå region \citep{wee16}. The population in the two
regions is large enough and socially diverse to enable studies of social
mortality inequalities. Poplink is based on linked parish records, allowing
us to reconstruct life biographies on people as long as they remained in
the region. The records are linked within but not between the regions. Data
from the regions are accessed for the period 1901--1950s.

The other large data-set comes from the Linnaeus database
\citep{malm10}. It is based on different linked population registers from
1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from censuses every fifth
year 1965--1990, with additional information on deaths from National Board of Health and Welfare. For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare. The construction of these data-sets for analysis is presented below.

Information on individuals are anonymized and the two databases are not
linked between each other. This stops us from following individuals present
in both databases throughout their lives, thus they are treated separately. This
also makes it impossible to add information on individuals in the Linnaeus
database from what we potentially could find in Poplink, for example family
background or occupations earlier in life.

In the data-set analysed here, all individuals 40 years and older and ever being resident in either
of the regions are included. The data file contains the variables on sex,
birth date, death date, first and last date of observation and type of
entrance/exit.

The identification of presence periods as well as the time-dependent variables require some additional comments.

\section{Models}

The proportional hazards model in a survival analysis context is used,
allowing for adjustment for civil status and rural/urban environment. Since
the central explanatory variable, {\sc Hisclass}, do not follow the
proportionality property, the analyses are stratified with respect to the
variable, and the main results are presented graphically.

The analyses are performed in the \R\ environment for statistical computing
and graphics \citep{cran}, especially using the package {\tt eha} \citep{gb12,eha}
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--64 and 65--99 years of age. The model controls for period, marital
status and whether the individual resides in the urban or rural part. There
are no serious signs of deviation from proportionality assumption, except
regarding {\sc Hisclass}.

We analyse both regions together but perform separate analyses for men and
women and for the age groups 40--64 and 65--89. The model in the analyses
include social class, marital status and urban/rural residence. Each decade
during the 20th century gets separate analyses, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period. We present the results for the period 2001--2013
separately. We suggest a heterogeneity index in order to investigate a
possible homogenisation of mortality between social classes.

The important explanatory variable, {\sc Hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\sc Hisclass} on survival is
non-proportional, that is, it varies with age. Thus, the main results are
presented \emph{graphically}. Note further that {\tt Hisclass} is a
\emph{time-varying} variable. This causes lack of information in the higher
ages around 1960--1980, because that kind of information is simply missing
to a large extent.

\section{Results, all causes of death}

@@

The analysis for the last period is based on a separate analysis with
income and education as explanatory variables.

@@

%%{\huge Här ska 'conclusion.tex' in !!!}
